---
vignette: >
  % \VignetteIndexEntry{BugSigDB Stats and Analysis}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output:
  html_document:
    mathjax: null  
    toc: true
    toc_depth: 2
    toc_float: true
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
library(bugsigdbr)
library(BugSigDBStats)
library(ComplexHeatmap)
library(ggpubr)
```

# Reading data

Get bulk export from bugsigdb.org:

```{r curationFile}
full.dat <- bugsigdbr::importBugSigDB(version = "devel", cache = FALSE)
dim(full.dat)
colnames(full.dat)
```

Stripping illformed entries:

```{r}
is.study <- grepl("^Study [0-9]+$", full.dat[["Study"]])
is.exp <- grepl("^Experiment [0-9]+$", full.dat[["Experiment"]])
full.dat <- full.dat[is.study & is.exp, ]
```

# Curation output

Number of papers and signatures curated:

```{r nrPapers}
pmids <- unique(full.dat[,"PMID"])
length(pmids)
nrow(full.dat)
```

## Publication date of the curated papers:

```{r pubDate}
pmids <- pmids[!is.na(pmids)]
pubyear <- pmid2pubyear(pmids)
head(cbind(pmids, pubyear))
```

```{r pubDate2}
tab <- table(pubyear)
tab <- tab[-length(tab)]
tab <- tab[order(as.integer(names(tab)))]
df <- data.frame(year = names(tab), papers = as.integer(tab))
ggbarplot(df, x = "year", y = "papers", 
          label = TRUE, fill = "steelblue",
          ggtheme = theme_bw())
``` 

Stripping empty signatures:

```{r} 
ind1 <- lengths(full.dat[["MetaPhlAn taxon names"]]) > 0
ind2 <- lengths(full.dat[["NCBI Taxonomy IDs"]]) > 0
dat <- full.dat[ind1 & ind2,]
nrow(dat)
```

Papers containing only empty UP and DOWN signatures (under curation?):
```{r}
setdiff(pmids, unique(dat[,"PMID"]))
```

## Progress over time:
```{r, fig.width=10, fig.height=10}
dat[,"Curated date"] <- as.character(lubridate::dmy(dat[,"Curated date"]))
plotProgressOverTime(dat)
plotProgressOverTime(dat, diff = TRUE)
```

Stratified by curator:
```{r curatorOutput, fig.width=10, fig.height=10}
npc <- stratifyByCurator(dat)
plotCuratorStats(dat, npc)
```

Number of complete and revised signatures:
```{r}
table(df[["State"]])
table(dat[,"Revision editor"])
```

# Study stats

## Study design

```{r}
spl <- split(dat[["Study"]], dat[["Study design"]])
sds <- lapply(spl, unique)
sort(lengths(sds), decreasing = FALSE)
```

# Experiment stats

Columns of the full dataset that describe experiments:

```{r}
# Experiment ID
exp.cols <- c("Study", "Experiment")

# Subjects
sub.cols <- c("Host species",    
              "Location of subjects", 
              "Body site",
              "Condition", 
              "Antibiotics exclusion",
              "Group 0 sample size",
              "Group 1 sample size")

# Lab analysis              
lab.cols <-  c("Sequencing type",
              "16S variable region",
              "Sequencing platform")

# Statistical analysis
stat.cols <-  c("Statistical test",
              "MHT correction",
              "Significance threshold")

# Alpha diversity
div.cols <- c("Pielou",
              "Shannon",
              "Chao1",
              "Simpson", 
              "Inverse Simpson",
              "Richness")
```

Restrict dataset to experiment information:

```{r}
exps <- dat[,c(exp.cols, sub.cols, lab.cols, stat.cols, div.cols)]
exps <- unique(exps)
```

## Subjects

Number of experiments for the top 10 categories for each subjects column:

```{r}
sub.tab <- lapply(sub.cols[1:5], tabCol, df = exps, n = 10)
names(sub.tab) <- sub.cols[1:5]
sub.tab
```

Proportions instead:

```{r}
sub.tab <- lapply(sub.cols[1:5], tabCol, df = exps, n = 10, perc = TRUE)
names(sub.tab) <- sub.cols[1:5]
sub.tab
```

Sample size:

```{r}
ssize <- apply(exps[,sub.cols[6:7]], 2, summary)
ssize
```

## Lab analysis

Number of experiments for the top 10 categories for each lab analysis column:

```{r}
lab.tab <- lapply(lab.cols, tabCol, df = exps, n = 10)
names(lab.tab) <- lab.cols
lab.tab
```

Proportions instead:

```{r}
lab.tab <- lapply(lab.cols, tabCol, df = exps, n = 10, perc = TRUE)
names(lab.tab) <- lab.cols
lab.tab
```

## Statistical analysis

Number of experiments for the top 10 categories for each statistical analysis column:

```{r}
stat.tab <- lapply(stat.cols, tabCol, df = exps, n = 10)
names(stat.tab) <- stat.cols
stat.tab
```

Proportions instead:

```{r}
stat.tab <- lapply(stat.cols, tabCol, df = exps, n = 10, perc = TRUE)
names(stat.tab) <- stat.cols
stat.tab
```

## Alpha diversity

Overall distribution:

```{r}
apply(exps[,div.cols], 2, table)
```

Correspondence of Shannon diversity and Richness:

```{r}
table(exps$Shannon, exps$Richness)
```

Conditions with consistently increased or decreased alpha diversity:

```{r}
tabDiv(exps, "Shannon", "Condition")
tabDiv(exps, "Shannon", "Condition", perc = TRUE)
tabDiv(exps, "Richness", "Condition")
tabDiv(exps, "Richness", "Condition", perc = TRUE)
```

Body sites with consistently increased or decreased alpha diversity:

```{r}
tabDiv(exps, "Shannon", "Body site")
tabDiv(exps, "Shannon", "Body site", perc = TRUE)
tabDiv(exps, "Richness", "Body site")
tabDiv(exps, "Richness", "Body site", perc = TRUE)
```

# Signature stats

```{r}
sigs <- bugsigdbr::getSignatures(dat, tax.id.type = "metaphlan")
```

## Unique microbes

Number unique microbes contained in the signatures:
```{r ubugs}
(nuniq <- length(unique(unlist(sigs))))
```

Development of unique microbes captured over time:
```{r}
plotUniqueMicrobesOverTime(dat)
```

## Microbe set size distribution

```{r setSize}
summary(lengths(sigs))
gghistogram(lengths(sigs), bins = 30, ylab = "number of signatures",
    xlab = "signature size", fill = "#00AFBB", ggtheme = theme_bw())
sum(lengths(sigs) > 4)
```

## Microbe co-occurrence

```{r coocc, fig.width = 10, fig.height = 10}
dat.feces <- subset(dat, `Body site` == "Feces")
cooc.mat <- microbeHeatmap(dat.feces, tax.level = "genus", anno = "genus")
antag.mat <- microbeHeatmap(dat.feces, tax.level = "genus", anno = "genus", antagonistic = TRUE)
```

Get the top 20 genera most frequently reported as differentially abundant:
 
```{r}
sigs.feces <- getSignatures(dat.feces, tax.id.type = "taxname", 
                            tax.level = "genus", exact.tax.level = FALSE) 
top20 <- sort(table(unlist(sigs.feces)), decreasing = TRUE)[1:20]
top20
```

Subset heatmaps to the top 20 genera most frequently reported as differentially abundant:

```{r}
all(names(top20) %in% rownames(cooc.mat))
cooc.mat <- cooc.mat[names(top20), names(top20)]
all(names(top20) %in% rownames(antag.mat))
antag.mat <- antag.mat[names(top20), names(top20)]
```

Distinguish by direction of abundance change (increased / decreased):

```{r}
# increased
sub.dat.feces <- subset(dat.feces, `Abundance in Group 1` == "increased")
sigs.feces.up <- getSignatures(sub.dat.feces, tax.id.type = "taxname", 
                               tax.level = "genus", exact.tax.level = FALSE) 
top20.up <- table(unlist(sigs.feces.up))[names(top20)]
top20.up

# decreased
sub.dat.feces <- subset(dat.feces, `Abundance in Group 1` == "decreased")
sigs.feces.down <- getSignatures(sub.dat.feces, tax.id.type = "taxname", 
                                 tax.level = "genus", exact.tax.level = FALSE) 
top20.down <- table(unlist(sigs.feces.down))[names(top20)]
top20.down
```

Plot the heatmap

```{r mheat, fig.width = 10, fig.height = 10}
# annotation
mat <- matrix(nc = 2, cbind(top20.up, top20.down))
bp <- ComplexHeatmap::anno_barplot(mat, gp = gpar(fill = c("#D55E00", "#0072B2"),
                                                  col = c("#D55E00", "#0072B2")),
                                   height = unit(2, "cm"))
banno <- ComplexHeatmap::HeatmapAnnotation(`Abundance in Group 1` = bp)

lgd_list <- list(
    Legend(labels = c("increased", "decreased"), 
           title = "Abundance in Group 1", 
           type = "grid",
           legend_gp = gpar(col = c("#D55E00", "#0072B2"), fill = c("#D55E00", "#0072B2"))))
                                            
# same direction
# lcm <- sweep(cooc.mat, 2, matrixStats::colMaxs(cooc.mat), FUN = "/")
# we need to dampen the maximum here a bit down,
# otherwise 100% self co-occurrence takes up a large fraction of the colorscale,
sec <- apply(cooc.mat, 2, function(x) sort(x, decreasing = TRUE)[2])
cooc.mat2 <- cooc.mat
for(i in 1:ncol(cooc.mat2)) cooc.mat2[i,i] <- min(cooc.mat2[i,i], 1.4 * sec[i])
lcm <- sweep(cooc.mat2, 2, matrixStats::colMaxs(cooc.mat2), FUN = "/")
col <- circlize::colorRamp2(c(0,1), c("#EEEEEE", "red"))
ht1 <- ComplexHeatmap::Heatmap(lcm,
                               col = col, 
                               name = "Relative frequency (top)",
                               cluster_columns = FALSE, 
                               row_km = 3, 
                               row_title = "same direction", 
                               column_names_rot = 45,
                               row_names_gp = gpar(fontsize = 8),
                               column_names_gp = gpar(fontsize = 8))

# opposite direction
acm <- sweep(antag.mat, 2, matrixStats::colMaxs(antag.mat), FUN = "/")
col <- circlize::colorRamp2(c(0,1), c("#EEEEEE", "blue"))
ht2 <- ComplexHeatmap::Heatmap(acm,
                               col = col, 
                               name = "Relative frequency (bottom)",
                               cluster_columns = FALSE, 
                               row_title = "opposite direction", 
                               row_km = 3, 
                               column_names_rot = 45,
                               row_names_gp = gpar(fontsize = 8),
                               column_names_gp = gpar(fontsize = 8))

# phylum
sfp <- bugsigdbr::getSignatures(dat.feces, tax.id.type = "metaphlan", 
                                tax.level = "genus", exact.tax.level = FALSE) 
sfp20 <- sort(table(unlist(sfp)), decreasing = TRUE)[1:20]
uanno <- bugsigdbr::extractTaxLevel(names(sfp20),
                                    tax.id.type = "taxname",
                                    tax.level = "phylum",
                                    exact.tax.level = FALSE) 
phyla.grid <- seq_along(unique(uanno))
panno <- ComplexHeatmap::HeatmapAnnotation(phylum = uanno)

uanno <- matrix(uanno, nrow = 1)
colnames(uanno) <- names(top20)
pcols <- c("#CC79A7", "#F0E442", "#009E73", "#56B4E9", "#E69F00")
uanno <- ComplexHeatmap::Heatmap(uanno, name = "Phylum",
                                 col = pcols[phyla.grid],
                                 cluster_columns = FALSE,
                                 column_names_rot = 45,
                                 column_names_gp = gpar(fontsize = 8))

# put everything together
ht_list <- ht1 %v% banno %v% ht2 %v% uanno
ComplexHeatmap::draw(ht_list, annotation_legend_list = lgd_list, merge_legend = TRUE)

decorate_annotation("Abundance in Group 1", {
    grid.text("# signatures", x = unit(-1, "cm"), rot = 90, just = "bottom", gp = gpar(fontsize = 8))
    grid.text("*", x = unit(2.45, "cm"), y = unit(1.2, "cm"))
    grid.text("*", x = unit(5.18, "cm"), y = unit(1, "cm"))
    grid.text("*", x = unit(6.55, "cm"), y = unit(0.95, "cm"))
    grid.text("*", x = unit(8.6, "cm"), y = unit(0.85, "cm"))
    grid.text("*", x = unit(10, "cm"), y = unit(0.7, "cm"))
    grid.text("*", x = unit(10.7, "cm"), y = unit(0.7, "cm"))
})
```

## Signature similarity

### Jaccard index
 
Inspect signature similarity for signatures from stomach samples based on Jaccard index:

```{r}
stomachsub <- subset(dat, `Body site` == "Stomach")
sigsub <- bugsigdbr::getSignatures(stomachsub)
pair.jsim <- calcJaccardSimilarity(sigsub)
```

Create a dendrogram of Jaccard dissimilarities (1.0 has no overlap, 0.0 are identical signatures). 

```{r, fig.height=15, fig.width=8}
jdist <- as.dist(1 - pair.jsim)
plot(hclust(jdist))
```
