---
vignette: >
  % \VignetteIndexEntry{Semantic similarity of BugSigDB signatures}
  % \VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %e, %Y")`'
output:
  html_document:
    mathjax: null   
---

# Setup
```{r setup, message=FALSE, warning=FALSE}
library(bugsigdbr)
library(ggtree)
library(ontologyIndex)
library(ontologySimilarity)
```

# Semantic similarity

Semantic similarity measures have been proposed for comparing concepts within an
ontology [Schlicker et al., 2006](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-7-302).
We therefore treat the NCBI taxonomy as an ontology, and compute semantic similarity
between signatures.

```{r} 
onto <- BugSigDBStats::getNcbiTaxonomyObo()
# onto <- ontologyIndex::get_ontology("http://purl.obolibrary.org/obo/ncbitaxon.obo")
```

```{r}
onto
head(onto$id)
```

One challenge here is that this needs quite some memory (kills your GHA):

```{r}
pryr::object_size(onto)
```

We add the corresponding ID prefix:

```{r, message = FALSE}
dat <- bugsigdbr::importBugSigDB()
dat.feces <- subset(dat, `Body site` == "feces")
ind <- lengths(dat.feces[["NCBI Taxonomy IDs"]]) > 5
dat.feces <- dat.feces[ind,]
sigs <- bugsigdbr::getSignatures(dat.feces, tax.id.type = "ncbi")
sigs <- lapply(sigs, function(s) paste0("NCBITaxon:", s))
```            

We remove taxa that are not in the NCBI Taxonomy:

```{r}
utax <- unique(unlist(sigs))
nt <- utax[!(utax %in% onto$id)]
sigs <- lapply(sigs, function(s) setdiff(s, nt))
```

Now, we compute pairwise semantic similarity for all signatures 
(which is reasonably fast btw):

```{r}
system.time(
    sim.mat <- ontologySimilarity::get_sim_grid(ontology = onto, term_sets = sigs)
)
dim(sim.mat)
sim.mat[1:5,1:5]
summary(as.vector(sim.mat))
```

Given the matrix of pairwise semantic similarity between signatures, we can also
compute the semantic similarity of a group of signatures by eg. taking the average
similarity between all pairs of signatures.

Here, we compute the semantic similarity of all colorectal cancer signatures in
the database for which abundance is increased in the cases. 

```{r}
ind <- !is.na(dat[["Condition"]]) &
       dat[["Condition"]] == "colorectal cancer" & 
       dat[["Body site"]] == "feces" & 
       dat[["Abundance in Group 1"]] == "increased"
sum(ind)
ontologySimilarity::get_sim(sim.mat, group = which(ind))
```

Furthermore, we can compute a p-value for assessing the statistical
significance of the similiarity of a group of signatures. The p-value is calculated
by random sampling of groups of the same size as `group`, and calculating how many
random groups have at least as great group similarity than does `group`.

```{r}
ontologySimilarity::get_sim_p(sim.mat, group = which(ind))
```

Cluster similarity and visualize signature:

```{r, warning = FALSE}
sm.grp <- sim.mat[ind,ind]
hc <- stats::hclust(stats::as.dist(1 - sm.grp), method = "ward.D")
clus <- stats::cutree(hc, 5)
d <- data.frame(label = names(clus), count = lengths(sigs[names(clus)]))
p <- enrichplot:::group_tree(hc, clus, d, group_color = NULL,
                             offset_tiplab = 0.35, label_format = 30, nWords = 0,
                             offset = -0.1, extend = 0.3, hilight = TRUE, fontsize = 4)
p + coord_cartesian(xlim = xlim) +
    ggnewscale::new_scale_colour() + 
    geom_tippoint(aes(size = count))
```

Chunk up entire BugSigDB by body site:

```{r}
ind <- !is.na(dat[["Body site"]]) & !grepl(",", dat[["Body site"]]) 
dat.bs <- dat[ind,]
sigs.bs <- bugsigdbr::getSignatures(dat.bs)
spl <- split(sigs.bs, dat.bs[["Body site"]])
spl <- lapply(spl, function(s) unique(unlist(s)))
ind <- lengths(spl) > 4
spl <- spl[ind]
sort(lengths(spl), decreasing = TRUE)
```

Calculate semantic similarity between body site chunked-up signatures:

```{r}
spl <- lapply(spl, function(s) paste0("NCBITaxon:", s))
utax <- unique(unlist(spl))
nt <- utax[!(utax %in% onto$id)]
spl <- lapply(spl, function(s) setdiff(s, nt))
sm.bs <- ontologySimilarity::get_sim_grid(ontology = onto, term_sets = spl)
sm.bs[1:5,1:5]
```

```{r, warning = FALSE}
hc <- stats::hclust(stats::as.dist(1 - sm.bs), method = "ward.D")
clus <- stats::cutree(hc, 5)
d <- data.frame(label = names(clus), count = lengths(spl[names(clus)]))
p <- enrichplot:::group_tree(hc, clus, d, group_color = NULL,
                             offset_tiplab = 0.35, label_format = 30, nWords = 0,
                             offset = -0.1, extend = 0.3, hilight = TRUE, fontsize = 4)
p + coord_cartesian(xlim = xlim) +
    ggnewscale::new_scale_colour() + 
    geom_tippoint(aes(size = count))
```

Take a closer look at nasal cavity signatures:

```{r}
ndat <- subset(dat, `Body site` == "nasal cavity")
nms <- sort(unique(unlist(ndat[["MetaPhlAn taxon names"]])))
dim(ndat)
unique(ndat$PMID)
```

Compare nasal cavity meta-signature to feces meta-signature:

```{r}
fdat <- subset(dat, `Body site` == "feces")
fms <- sort(unique(unlist(fdat[["MetaPhlAn taxon names"]])))
table(nms %in% fms)
```

Check whether remaining nasal cavity taxa are in feces meta-signature at family level:

```{r}
nnms <- nms[!(nms %in% fms)]
nnms.fam <- bugsigdbr::extractTaxLevel(nnms, tax.level = "family", exact.tax.level = FALSE) 
fms.fam <- bugsigdbr::extractTaxLevel(fms, tax.level = "family", exact.tax.level = FALSE)
table(nnms.fam %in% fms.fam)
```
